---
title: "HW2 STA521 Fall18"
author: 'Shuai Yuan, sy144, Kolin96'
date: "Due September 23, 2018 5pm"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Backgound Reading

Readings: Chapters 3-4 in Weisberg Applied Linear Regression


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(knitr)
library(GGally)
library(dplyr)
```

<!-- This exercise involves the UN data set from `alr3` package. Install `alr3` and the `car` packages and load the data to answer the following questions adding your code in the code chunks.  Please add appropriate code to the chunks to suppress messages and warnings as needed once you are sure the code is working properly and remove instructions if no longer needed. Figures should have informative captions. Please switch the output to pdf for your final version to upload to Sakai. **Remove these instructions for final submission** -->

## Exploratory Data Analysis

<!-- 0.  Preliminary read in the data.  After testing, modify the code chunk so that output, messages and warnings are suppressed.  *Exclude text from final* -->

```{r data, include=FALSE}
library(alr3)
data(UN3, package="alr3")
help(UN3) 
library(car)
```

1. Create a summary of the data.  How many variables have missing data?  Which are quantitative and which are qualtitative?

```{r}
summary(UN3)
```
Six variables have missing data: `ModernC`, `Change`, `PPgdp`, `Frate`, `Pop` and `Fertility`.

Quantitative: `ModernC`, `Change`, `PPgdp`, `Frate`, `Pop`, `Fertility`, `Purban`  

Qualitative: (none)

2. What is the mean and standard deviation of each quantitative predictor?  Provide in a nicely formatted table.

```{r}
mean_result <- sapply(UN3, mean, na.rm=TRUE)
std_result <- sapply(UN3, sd, na.rm=TRUE)
kable(cbind(mean_result, std_result), col.names = c("mean", "std."), digits = 4)
```

3. Investigate the predictors graphically, using scatterplots or other tools of your choice. Create some plots
highlighting the relationships among the predictors. Comment
on your findings regarding trying to predict `ModernC` from the other variables.  Are there potential outliers, nonlinear relationships or transformations that appear to be needed based on your graphical EDA?

```{r}
gp = ggpairs(UN3, titl="The ggpairs plot for UN3")
print(gp, progress=FALSE)
```

If we try to predict `ModernC`, it may be better to use the predictors `Change`, `Frate`, `Fertility` and `Purban`, since they are linear correlated.

Potential outliers: the `Pop` variable has two extremely large data point.

Nonlinear relationships: `PPgdp` seems to grow exponentially as `ModernC` increases. `Frate` also has some noised nonlinear relationship.

Transformation needed: `PPgdp` should be transformed to log scale.

## Model Fitting

4.  Use the `lm()` function to perform a multiple linear regression with `ModernC` as the response and all other variables as the predictors, using the formula `ModernC ~ .`, where the `.` includes all remaining variables in the dataframe.  Create  diagnostic residual plot from the linear model object and comment on results regarding assumptions.  How many observations are used in your model fitting?

```{r}
mymodel = lm(ModernC ~ ., data=UN3)
summary(mymodel)
# diagnostic residual plot
plot(mymodel)
```

Comments: The linear model works fairly well, since the R-squared values are high (around 0.6) and the p-value is low (<2.2e-16). There are noises, but the residuals are generally independent of fitted values and have mean 0. For the Normal Q-Q plot, the higher quantiles can't match perfectly, but it is still not bad. Cook's distances are all below 0.5; China and Indea are potential outliers.

How many observations are used: 125, since the other 85 observations are deleted due to missingness.

5. Examine added variable plots `car::avPlot` or `car::avPlots`  for your model above. Are there any plots that suggest that transformations are needed for any of the terms in the model? Describe. Is it likely that any of the localities are influential for any of the terms?  Which localities?  Which terms?  

```{r}
car::avPlots(mymodel)
```

Transformations that are needed: The `Pop | others` plot shows that we need to transform `pop` to log scale, because all data points are now concentrated near 0.

Influential localities for each term are all shown in the figures:

`Change`: Cook islands, Kuwaito, Poland, Azerbaijan;

`PPgdp`: Switzerland, Norway, Azerbaijan;

`Frate`: Yeman, Burundi, Poland, Azerbaijan;

`Pop`: China, Indea, Azerbaijan;

`Fertility`: Thailand, Niger, Poland, Azerbaijan;

`Purban`: Thailand, Sri.Lanka, Poland, Azerbaijan.

Note that Azerbaijan appears in all the lists.

6.  Using the Box-Tidwell  `car::boxTidwell` or graphical methods find appropriate transformations of the predictor variables to be used as predictors in the linear model.  If any predictors are negative, you may need to transform so that they are non-negative.  Describe your method and  the resulting transformations.


```{r}
UN3_nonnega = UN3
UN3_nonnega['Change'] = UN3_nonnega['Change'] - min(UN3['Change'], na.rm=TRUE) + 1
car::boxTidwell(ModernC ~  Pop, ~ PPgdp + Fertility + Change + Frate + Purban, data=UN3_nonnega, na.action = na.omit)
```

We first transform `Change` by subtracting its minimum value to make it nonnegative. Then, we transform `Pop`, the MLE of lambda computed by `car::boxTidwell` is 0.63309.

7. Given the selected transformations of the predictors, select a transformation of the response using `MASS::boxcox` or `car::boxCox` and justify.


```{r}
mymodel = lm(ModernC ~ ., data=UN3_nonnega)
lambda = car::boxCox(mymodel)
```

By outputing car::boxCox(mymodel), we may find the best lambda is around 0.7879.

8.  Fit the regression using the transformed variables.  Provide residual plots and added variables plots and comment.  If you feel that you need additional transformations of either the response or predictors, repeat any steps until you feel satisfied.

```{r}
UN3_nonnega = UN3_nonnega %>% mutate(logPop = log(Pop))
mymodel = lm(ModernC ~ PPgdp + Fertility + Change + Frate + Purban + logPop, data=UN3_nonnega)
summary(mymodel)
plot(mymodel)
car::avPlots(mymodel)
```

We also applies log transformation to `Pop`. The residual plots show that residuals are very small (close to mean 0) and independent to fitted values. The Cook's distance are all very small. In the added variable plot, the `logPop` plot is much more balanced than the original `Pop` plot. The new model works better than our previous model.

9. Start by finding the best transformation of the response and then find transformations of the predictors.  Do you end up with a different model than in 8?


```{r}

```

Yes, we would get different models, but these models are not necessarily much better than our previous model, because the original response variable actually has strong linear correalations with many predictors. If we first transform the response variable by log, poly, etc., we still need to transform many predictors accordingly, making the model very messy. Actually, our previous model is good enough.

10.  Are there any outliers or influential points in the data?  Explain.  If so, refit the model after removing any outliers and comment on residual plots.

```{r}
abs.ti = abs(rstudent(mymodel))
pval= 2*(1- pt(max(abs.ti), mymodel$df - 1))
```

We can use student t test with Bonferroni Correction to determine whether there are outliers. The p-value for the observation with largest studentized residual is much lower than 0.05/125, so we claim that there is no outlier. There should be also no influential points, since in the former plots, all Cook's distances are low.

## Summary of Results

11. For your final model, provide summaries of coefficients with 95% confidence intervals in a nice table with interpretations of each coefficient.  These should be in terms of the original units! 


```{r}

```


12. Provide a paragraph summarizing your final model  and findings suitable for the US envoy to the UN after adjusting for outliers or influential points.   You should provide a justification for any case deletions in your final model


```{r}

```


## Methodology

    
13. Prove that the intercept in the added variable scatter plot will always be zero.  _Hint:  use the fact that if $H$ is the project matrix which contains a column of ones, then $1_n^T (I - H) = 0$.  Use this to show that the sample mean of residuals will always be zero if there is an intercept._


14. For multiple regression with more than 2 predictors, say a full model given by `Y ~ X1 + X2 + ... Xp`   we create the added variable plot for variable `j` by regressing `Y` on all of the `X`'s except `Xj` to form `e_Y` and then regressing `Xj` on all of the other X's to form `e_X`.  Confirm that the slope in a manually constructed added variable plot for one of the predictors  in Ex. 10 is the same as the estimate from your model. 
